### 2025-01-05 2024-5-13/sections/algorithms.tex
Final version of revision

At first, we show how to construct a CEFA from a regex where counting operators may occur. \hdh{Our approach largely follows Thompson's construction~\cite{Thompson1968} but we avoid using the $\epsilon$-transitions. Similar ideas of representing counting operators as counters in automata have been used in e.g. \cite{redos_lenka}. }
%\hdh{and we remove $\epsilon$-transitions on the fly (See union at Section \ref{sec:automaton} and construction of counting operators below)}. 
%\hdh{Some other algorithms to construct NFA from regular expressions are also available, for instance, Glushkov’s construction~\cite{Gluskov61} and its variants~\cite{redos_lenka,HS+23,GKM23}.}

@@

\medskip
\noindent
\emph{Step 3 (Computing an LIA formula).} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then the $\cefadec$ problem is reduced to the satisfiability of the following LIA formula, 
%
\begin{equation}\label{eq:parikh_image}
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\end{equation}

@@

\medskip
\noindent
\emph{Step 4 (Solving the LIA formula).}
We utilize some off-the-shelf SMT solver (e.g. Z3) to solve the satisfiability of the LIA formula in Equation~(\ref{eq:parikh_image}). If the solver reports ``sat'' and returns a solution $\amodel$, then we go to Step 5. Otherwise, the $\cefadec$ problem has no solution.

\hdh{
\medskip
\noindent
\emph{Step 5 (Constructing a solution).}
%For each $i \in [n]$, 
Let $\amodel$ be a solution of the LIA formula in Equation~(\ref{eq:parikh_image}). We want to construct from $\amodel$ and the CEFAs $\cB_i$ for $i \in [n]$, a solution $\amodel'$ of the $\cefadec$ problem. The construction of $\amodel'$ is obtained by constructing $\amodel'(x_i)$ for each $i \in [n]$. 
Let $i \in [n]$. We describe how to construct $\amodel'(x_i)$. 
The free variables $\fZ_{\costset_{\cC_i}}$ correspond to the number of occurrences of characters in $\costset_{\cC_i}$, i.e. the alphabet of the NFA $\cC_i$. Moreover, in $\cC_i$, the characters are just the integer vectors and the characters in the original CEFAs $\cB_i$ used to construct $\cC_i$ are ignored. 
The construction of $\amodel'(x_i)$ is actually to solve the reachability problems in a finite transition system where 
\begin{itemize}
\item the nodes are of the form $(q, (\nvec_{i, \myvec{v}})_{\myvec{v} \in \costset_{\cC_i}})$, where $\nvec_{i, \myvec{v}}$ denotes the number of the remaining occurrences of the integer vector $\myvec{v}$ in $\cB_i$, 
%
\item the transitions are of the form 
$$((q, (\nvec_{i, \myvec{v}})_{\myvec{v} \in \costset_{\cC_i}}), (q, a, q', \myvec{u}), (q', (\nvec'_{i, \myvec{v}})_{\myvec{v} \in \costset_{\cC_i}}))$$
such that $(q, a, q', \myvec{u})$ is a transition in $\cB_i$, $\nvec'_{i, \myvec{u}} = \nvec_{i, \myvec{u}}-1$, moreover, $\nvec'_{i, \myvec{v}} = \nvec_{i, \myvec{v}}$ for every $\myvec{v} \neq \myvec{u}$. 
\end{itemize}
Moreover, the initial and final state of the reachability problem are set to be $(q_0, (\amodel(\anivarz_{i, \myvec{v}}))_{\myvec{v} \in \costset_{\cC_i}})$ and $(q_f, (0,\cdots, 0))$ for some initial state $q_0$ and final state $q_f$ in $\cB_i$. 
If a path from the initial state to the final state is found, then the path gives us a desired string $\amodel'(x_i)$.
}
%Note that the transition system contains exponentially many nodes. As a result, the searching process takes \emph{doubly exponential time} in the worst case, since there are at most doubly exponentially many simple paths. To alleviate the path-explosion problem, in the sequel, we propose two techniques to reduce the number of transitions of the transition system.

@@

\hdh{
\subsection{Complexity Analysis of the $\reclsat$ Problem}
The time complexity of regular expression equivalence with complement has been established as nonelementary~\cite{complementary_complexity_1974}. This directly implies that the $\reclsat$ problem exhibits a nonelementary lower bound. We now turn our attention to analyzing the complexity of the $\reclsat$ problem in the absence of complement operations (no occurrence of $\Bar{e}$ and $e_1 \backslash e_2$).

\begin{lemma}\label{lemma:regex2cefa}
    Register-representable regex without complement operation can be converted to CEFA in EXPSPACE.
\end{lemma}

\begin{proof}
    We prove this lemma through structural induction on regular expression $e$. For the base case—where the regular expression consists of a single character, $\epsilon$, or $\emptyset$—the lemma holds trivially.
    For the inductive cases, assuming the lemma holds for regular expressions $e_1$ and $e_2$:
    When $e = e_1\concat e_2$ or $e = e_1+e_2$, the concatenation or union operations described in Section \ref{sec:automaton} produce a CEFA of size $O(2^{|e_1|} + 2^{|e_2|}) = O(2^{\max(|e_1|, |e_2|)}) = O(2^{|e|})$.
    For $e = e_1\cap e_2$, the intersection operation detailed in Section \ref{sec:automaton} yields a CEFA of size $O(2^{|e_1|}\cdot 2^{|e_2|})=O(2^{|e_1| + |e_2|})=O(2^{|e|})$.
    In the case of $e=e_1^{\{m,n\}}$, our construction in Section \ref{subsec:regex2cefa} avoids direct unfolding of counting operators, resulting in a CEFA of size $O(2^{|e_1|}) = O(2^{|e|})$.
    Finally, for $e=(e_1)$ and $e=e^*$, our application of Thompson's construction~\cite{Thompson1968} produces a CEFA of size $O(2^{|e_1|}) = O(2^{|e|})$.
\end{proof}

\begin{theorem}\label{theorem_pspace}
    If all regexes are register-representable and complement operations are excluded, $\reclsat$ can be solved in EXPSPACE.
\end{theorem}
\begin{proof}
    Our decision procedure for $\reclsat$ with only register-representable regexes can be decomposed into two steps: (1) constructing CEFAs from regular membership and length function, (2) solving $\cefadec$.
    Based on Lemma \ref{lemma:regex2cefa}, our construction yields a CEFA of exponential size for each register-representable regex without complement operation. Furthermore, for each string variable's length function, a CEFA of constant size is constructed in Section \ref{subsec:regex2cefa}.
    Consequently, constructing CEFAs in the decision procedure outlined in Section \ref{subsec:regex2cefa} requires nondeterministic exponential size. After constructing CEFAs, the $\reclsat$ problem is transferred to the $\cefadec$ problem. By Proposition \ref{prop-cefadec}, we can demonstrate that the decision procedure likewise requires nondeterministic exponential size. Therefore, the complexity of the decision procedure for solving the restricted $\reclsat$ problem is NEXPSPACE, which yields EXPSPACE by Savitch's theorem~\cite{complexity-book}.
\end{proof}

\begin{theorem}
    $\reclsat$ can be solved in 3-EXPTIME when complement operations are excluded.
\end{theorem}

\begin{proof}
    Our decision procedure for $\reclsat$ with non-register-representable regexes can be decomposed into three steps: (1) unfolding non-register-representable occurrences of counting operators, (2) constructing CEFAs from regular membership and length function, (3) solving $\cefadec$.
    From the decision procedure in Section \ref{subsec:regex2cefa}, unfolding non-register-representable occurrences of counting operators leads to an exponential growth in expression length. For each non-register-representable regular expression of length $l$, the unfolding process generates a register-representable regular expression of length $O(m^n\cdot l)=O((2^{|l|})^{|l|}\cdot l) = O(2^{|l|})$, where $m$ represents the maximum counting bound and $n$ denotes the maximum depth of nested counting operators. By Theorem \ref{theorem_pspace} and the fact that EXPSPACE $\subseteq$ 2-EXPTIME \cite{complexity-book}, the complexity of the decision procedure for solving the $\reclsat$ problem without complement operations is 3-EXPTIME.
\end{proof}
}

>>>>

At first, we show how to construct a CEFA from a regex where counting operators may occur. \hdh{Our approach largely follows Thompson's construction~\cite{Thompson1968} but we avoid using the $\epsilon$-transitions. Similar ideas of representing counting operators as counters in automata have appeared in e.g. \cite{redos_lenka}. }
%\hdh{and we remove $\epsilon$-transitions on the fly (See union at Section \ref{sec:automaton} and construction of counting operators below)}. 
%\hdh{Some other algorithms to construct NFA from regular expressions are also available, for instance, Glushkov’s construction~\cite{Gluskov61} and its variants~\cite{redos_lenka,HS+23,GKM23}.}

@@

\medskip
\noindent
\emph{Step 3 (Computing an LIA formula).} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then the $\cefadec$ problem is reduced to the satisfiability of the following LIA formula,  \qed
%
\begin{equation}\label{eq:parikh_image}
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\end{equation}

@@

\medskip
\noindent
\emph{Step 4 (Solving the LIA formula).}
We utilize some off-the-shelf SMT solver (e.g. Z3) to solve the satisfiability of the LIA formula in Equation~(\ref{eq:parikh_image}). If the solver reports ``sat'' and returns a solution $\amodel$, then we go to Step 5. Otherwise, the $\cefadec$ problem has no solution.  \qed

\hdh{
\medskip
\noindent
\emph{Step 5 (Constructing a solution).}
%For each $i \in [n]$, 
Let $\amodel$ be a solution of the LIA formula in Equation~(\ref{eq:parikh_image}). We want to construct from $\amodel$ and the CEFAs $\cB_i$ for $i \in [n]$, a solution $\amodel'$ of the $\cefadec$ problem. The construction of $\amodel'$ is obtained by constructing $\amodel'(x_i)$ for each $i \in [n]$. 
Let $i \in [n]$. We describe how to construct $\amodel'(x_i)$. 
The free variables $\fZ_{\costset_{\cC_i}}$ correspond to the number of occurrences of characters in $\costset_{\cC_i}$, i.e. the alphabet of the NFA $\cC_i$. Moreover, in $\cC_i$, the characters are just the integer vectors and the characters in the original CEFAs $\cB_i$ used to construct $\cC_i$ are ignored. 
The construction of $\amodel'(x_i)$ is actually to solve the reachability problem in a finite transition system where 
\begin{itemize}
\item the nodes are of the form $(q, (\nvec_{i, \myvec{v}})_{\myvec{v} \in \costset_{\cC_i}})$, where $\nvec_{i, \myvec{v}}$ denotes the number of the remaining occurrences of the integer vector $\myvec{v}$ in $\cB_i$, 
%
\item the transitions are of the form 
$$((q, (\nvec_{i, \myvec{v}})_{\myvec{v} \in \costset_{\cC_i}}), (q, a, q', \myvec{u}), (q', (\nvec'_{i, \myvec{v}})_{\myvec{v} \in \costset_{\cC_i}}))$$
such that $(q, a, q', \myvec{u})$ is a transition in $\cB_i$, $\nvec'_{i, \myvec{u}} = \nvec_{i, \myvec{u}}-1$, moreover, $\nvec'_{i, \myvec{v}} = \nvec_{i, \myvec{v}}$ for every $\myvec{v} \neq \myvec{u}$. 
\end{itemize}
Moreover, the initial and final state of the reachability problem are set to be $(q_0, (\amodel(\anivarz_{i, \myvec{v}}))_{\myvec{v} \in \costset_{\cC_i}})$ and $(q_f, (0,\cdots, 0))$ for some initial state $q_0$ and final state $q_f$ in $\cB_i$. 
If a path from the initial state to the final state is found, then the path gives us a desired string $\amodel'(x_i)$. \qed
}
%Note that the transition system contains exponentially many nodes. As a result, the searching process takes \emph{doubly exponential time} in the worst case, since there are at most doubly exponentially many simple paths. To alleviate the path-explosion problem, in the sequel, we propose two techniques to reduce the number of transitions of the transition system.

@@

### 2025-01-05 2024-5-13/sections/cefa.tex
Final version of revision

The \emph{union} of $\aut_1$ and $\aut_2$ is denoted by $\aut_1 \cup \aut_2$. Two fresh auxiliary registers say $r'_1,r'_2 \not \in R_1 \cup R_2$, are introduced so that the accepting condition knows whether a run is from $\aut_1$ (or $\aut_2$).\hdh{This is mainly because the same accepting condition is applied to all final states, in other words, the accepting condition does not distinguish between final states.} Specifically, $\aut_1 \cup \aut_2 = (R', Q', \Sigma, \delta', I', F', \alpha')$ where 
\begin{itemize}
\item $R' = R_1 \cup R_2 \cup \{r'_1, r'_2\}$, $Q' = Q_1 \cup Q_2 \cup \{q'_0\}$ with $q'_0 \not \in Q_1 \cup Q_2$, $I' = \{q'_0\}$, 
%
\item $\delta'$ is the union of four transitions sets:
\begin{itemize}
  \item $\{(q'_0, a, q'_1, (\myvec{v}_1, \myvec{0}, 1,0)) \mid \exists q_1 \in I_1.\ (q_1, a, q'_1, \myvec{v}_1) \in \delta_1\}$
  \item $\{(q'_0, a, q'_2, (\myvec{0}, \myvec{v}_2, 0, 1)) \mid \exists q_2 \in I_2.\ (q_2, a, q'_2, \myvec{v}_2) \in \delta_2\}$
  \item $\{(q_1, a, q'_1, (\myvec{v}_1, \myvec{0}, 0, 0)) \mid q_1\not\in I_1.\ (q_1, a, q'_1, \myvec{v}_1) \in \delta_1\}$
  \item $\{(q_2, a, q'_2, (\myvec{0}, \myvec{v}_2, 0, 0)) \mid q_2\not\in I_2.\ (q_2, a, q'_2, \myvec{v}_2) \in \delta_2\}$
\end{itemize}
 where $(\myvec{v}_1, \myvec{0}, 1,0)$ is a vector that updates $R_1$ by $\myvec{v}_1$, updates $R_2$ by $\myvec{0}$, and updates $r'_1,r'_2$ by $1,0$ respectively. Similarly for $(\myvec{0}, \myvec{v}_2, 0, 1)$, and so on.

\item $F'$ and $\alpha'$ are defined as follows, 
\begin{itemize}
\item if $(\epsilon; \myvec{0})$ belongs to $\lan(\aut_1)$ or $\lan(\aut_2)$, i.e., one of the two automata accepts the empty string $\epsilon$, then $F' = F_1 \cup F_2 \cup \{q'_0\}$ and $\alpha' = (r'_1 = 0 \wedge r'_2 = 0) \vee (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$, 
\item otherwise, $F'=F_1 \cup F_2$ and $\alpha' = (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$.
\end{itemize}
\end{itemize}
From the construction, we know that 
$$
\Lang(\aut_1 \cup \aut_2) = 
\left\{(w; \myvec{v}) \left \vert 
\begin{array}{l}
%\myvec{v} \in \Int^{R'}, 
(w; \prj_{R_1}(\myvec{v})) \in \Lang(\aut_1) \mbox{ and } \prj_{r_1'}(\myvec{v}) = 1, \mbox{ or } \\
(w; \prj_{R_2}(\myvec{v})) \in \Lang(\aut_2) \mbox{ and } \prj_{r_2'}(\myvec{v}) = 1, \mbox{ or } \\
(w; \myvec{v}) = (\epsilon, \myvec{0}) \mbox{ if } \aut_1 \mbox{ or } \aut_2 \mbox{ accepts } \epsilon
\end{array}
\right.
\right\}.
$$ 
Intuitively, $\aut_1 \cup \aut_2$ accepts the words that are accepted by one of the $\aut_1$ and $\aut_2$ and outputs the costs of the corresponding automaton.

>>>>

The \emph{union} of $\aut_1$ and $\aut_2$ is denoted by $\aut_1 \cup \aut_2$. Two fresh auxiliary registers say $r'_1,r'_2 \not \in R_1 \cup R_2$, are introduced so that the accepting condition knows whether a run is from $\aut_1$ (or $\aut_2$). \hdh{Note that the fresh auxiliary registers are introduced to distinguish whether $\aut_1$ or $\aut_2$ accepts and weave the two accepting conditions of $\aut_1$ and $\aut_2$ into one accepting condition.}   
%is mainly because the same accepting condition is applied to all final states, in other words, the accepting condition does not distinguish between final states.} 
Specifically, $\aut_1 \cup \aut_2 = (R', Q', \Sigma, \delta', I', F', \alpha')$ where 
\begin{itemize}
\item $R' = R_1 \cup R_2 \cup \{r'_1, r'_2\}$, $Q' = Q_1 \cup Q_2 \cup \{q'_0\}$ with $q'_0 \not \in Q_1 \cup Q_2$, $I' = \{q'_0\}$, 
%
\item $\delta'$ is the union of four transitions sets:
\begin{itemize}
  \item $\{(q'_0, a, q'_1, (\myvec{v}_1, \myvec{0}, 1,0)) \mid \exists q_1 \in I_1.\ (q_1, a, q'_1, \myvec{v}_1) \in \delta_1\}$
  \item $\{(q'_0, a, q'_2, (\myvec{0}, \myvec{v}_2, 0, 1)) \mid \exists q_2 \in I_2.\ (q_2, a, q'_2, \myvec{v}_2) \in \delta_2\}$
  \item $\{(q_1, a, q'_1, (\myvec{v}_1, \myvec{0}, 0, 0)) \mid q_1\not\in I_1.\ (q_1, a, q'_1, \myvec{v}_1) \in \delta_1\}$
  \item $\{(q_2, a, q'_2, (\myvec{0}, \myvec{v}_2, 0, 0)) \mid q_2\not\in I_2.\ (q_2, a, q'_2, \myvec{v}_2) \in \delta_2\}$
\end{itemize}
 where $(\myvec{v}_1, \myvec{0}, 1,0)$ is a vector that updates $R_1$ by $\myvec{v}_1$, updates $R_2$ by $\myvec{0}$, and updates $r'_1,r'_2$ by $1,0$ respectively. Similarly for $(\myvec{0}, \myvec{v}_2, 0, 1)$, and so on.

\item $F'$ and $\alpha'$ are defined as follows, 
\begin{itemize}
\item if $(\epsilon; \myvec{0})$ belongs to $\lan(\aut_1)$ or $\lan(\aut_2)$, i.e., one of the two automata accepts the empty string $\epsilon$, then $F' = F_1 \cup F_2 \cup \{q'_0\}$ and $\alpha' = (r'_1 = 0 \wedge r'_2 = 0) \vee (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$, 
\item otherwise, $F'=F_1 \cup F_2$ and $\alpha' = (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$.
\end{itemize}
\end{itemize}
From the construction, we know that 
$$
\Lang(\aut_1 \cup \aut_2) = 
\left\{(w; \myvec{v}) \left \vert 
\begin{array}{l}
%\myvec{v} \in \Int^{R'}, 
(w; \prj_{R_1}(\myvec{v})) \in \Lang(\aut_1) \mbox{ and } \prj_{r_1'}(\myvec{v}) = 1, \mbox{ or } \\
(w; \prj_{R_2}(\myvec{v})) \in \Lang(\aut_2) \mbox{ and } \prj_{r_2'}(\myvec{v}) = 1, \mbox{ or } \\
(w; \myvec{v}) = (\epsilon, \myvec{0}) \mbox{ if } \aut_1 \mbox{ or } \aut_2 \mbox{ accepts } \epsilon
\end{array}
\right.
\right\}.
$$ 
Intuitively, $\aut_1 \cup \aut_2$ accepts the words that are accepted by one of the $\aut_1$ and $\aut_2$ and outputs the costs of the corresponding automaton. Note that an integer vector $\myvec{v}_1$ in $\aut_1$, is copied at most twice in $\aut_1 \cup \aut_2$, namely, $(\myvec{v}_1, \myvec{0}, 0,0)$ and $(\myvec{v}_1, \myvec{0}, 1,0)$. Simiarly for the integer vectors in $\aut_2$.
